---
title: "Evaluating Consequences of Storm Events"
author: "Jesus Dacuma"
date: "October 25, 2015"
output: html_document
---

## Synopsis

## Data Processing
Storm data can be donwloaded from the NOAA Storm Database. The `dplyr` library is used to convert the data frame to a tbl_df.
```{r, cache=TRUE, warning=FALSE, message=FALSE}
download.file("http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", "StormData.csv.bz2")
stormData <- read.csv("StormData.csv.bz2")

library(dplyr)
storm.tbl <- tbl_df(stormData)
```

The loaded dataset contains `r nrow(storm.tbl)` rows, and since the data contains events since 1955, there are many inconsistencies with the entries. To correctly evaluate the consequences of weather events, the first step is to clean up the dataset. 

### Cleaning Event Types (EVTYPE)

The raw data contains more than 900 levels of events, but many of these are the same and just vary in terminology (FROST vs. FREEZE), abbreviations (TSTM vs. THUNDERSTORM), case (RAIN vs. Rain), typos and other.

The process used to clean the data set was dynamic that involved looking at the most common event types (`head(storm.tbl %>% count(EVTYPE, sort = TRUE), 20)`) and searching for similar event names using regular expressions (`unique(storm.tbl$EVTYPE[grep("event", storm.tbl$EVTYPE, ignore.case = TRUE)])`). 

```{r}
# Combine similar names of event types
storm.tbl$EVTYPE[grep("TSTM", storm.tbl$EVTYPE, ignore.case = TRUE)] <- "TSTM WIND"
storm.tbl$EVTYPE[grep("THUNDER", storm.tbl$EVTYPE, ignore.case = TRUE)] <- "TSTM WIND"
storm.tbl$EVTYPE[grep("HAIL", storm.tbl$EVTYPE, ignore.case = TRUE)] <- "HAIL"
storm.tbl$EVTYPE[grep("TORNADO", storm.tbl$EVTYPE, ignore.case = TRUE)] <- "TORNADO"
storm.tbl$EVTYPE[grep("FLOOD", storm.tbl$EVTYPE, ignore.case = TRUE)] <- "FLOOD"
storm.tbl$EVTYPE[grep("HIGH WIND", storm.tbl$EVTYPE, ignore.case = TRUE)] <- "HIGH WIND"
storm.tbl$EVTYPE[grep("SNOW", storm.tbl$EVTYPE, ignore.case = TRUE)] <- "SNOW"
storm.tbl$EVTYPE[grep("RAIN", storm.tbl$EVTYPE, ignore.case = TRUE)] <- "RAIN"
storm.tbl$EVTYPE[grep("LIGHTNING", storm.tbl$EVTYPE, ignore.case = TRUE)] <- "LIGHTNING"
storm.tbl$EVTYPE[grep("WINTER", storm.tbl$EVTYPE, ignore.case = TRUE)] <- "WINTER STORM"
storm.tbl$EVTYPE[grep("SPOUT", storm.tbl$EVTYPE, ignore.case = TRUE)] <- "WATERSPOUT"
storm.tbl$EVTYPE[grep("FUNNEL", storm.tbl$EVTYPE, ignore.case = TRUE)] <- "FUNNEL CLOUD"
storm.tbl$EVTYPE[grep("CLOUD", storm.tbl$EVTYPE, ignore.case = TRUE)] <- "FUNNEL CLOUD"
storm.tbl$EVTYPE[grep("STRONG WIND", storm.tbl$EVTYPE, ignore.case = TRUE)] <- "STRONG WIND"
storm.tbl$EVTYPE[grep("URBAN", storm.tbl$EVTYPE, ignore.case = TRUE)] <- "URBAN/SML STREAM FLD"
storm.tbl$EVTYPE[grep("STREAM", storm.tbl$EVTYPE, ignore.case = TRUE)] <- "URBAN/SML STREAM FLD"
storm.tbl$EVTYPE[grep("FROST", storm.tbl$EVTYPE, ignore.case = TRUE)] <- "FROST/FREEZE"
storm.tbl$EVTYPE[grep("FREEZE", storm.tbl$EVTYPE, ignore.case = TRUE)] <- "FROST/FREEZE"
storm.tbl$EVTYPE[grep("FIRE", storm.tbl$EVTYPE, ignore.case = TRUE)] <- "WILDFIRE"
storm.tbl$EVTYPE[grep("BLIZZARD", storm.tbl$EVTYPE, ignore.case = TRUE)] <- "BLIZZARD"
storm.tbl$EVTYPE[grep("MICROBURST", storm.tbl$EVTYPE, ignore.case = TRUE)] <- "MICROBURST"
storm.tbl$EVTYPE[grep("DROUGHT", storm.tbl$EVTYPE, ignore.case = TRUE)] <- "DROUGHT/EXCESSIVE HEAT"
storm.tbl$EVTYPE[grep("HEAT", storm.tbl$EVTYPE, ignore.case = TRUE)] <- "DROUGHT/EXCESSIVE HEAT"
storm.tbl$EVTYPE[grep("DRY", storm.tbl$EVTYPE, ignore.case = TRUE)] <- "DROUGHT/EXCESSIVE HEAT"
storm.tbl$EVTYPE[grep("WET", storm.tbl$EVTYPE, ignore.case = TRUE)] <- "UNSEASONABLY WET"
storm.tbl$EVTYPE[grep("COLD", storm.tbl$EVTYPE, ignore.case = TRUE)] <- "EXTREME COLD/WIND CHILL"
storm.tbl$EVTYPE[grep("CHILL", storm.tbl$EVTYPE, ignore.case = TRUE)] <- "EXTREME COLD/WIND CHILL"
storm.tbl$EVTYPE[grep("ICE STORM", storm.tbl$EVTYPE, ignore.case = TRUE)] <- "ICE STORM"
storm.tbl$EVTYPE[grep("FOG", storm.tbl$EVTYPE, ignore.case = TRUE)] <- "DENSE FOG"
storm.tbl$EVTYPE[grep("SURF", storm.tbl$EVTYPE, ignore.case = TRUE)] <- "RIP CURRENTS/HEAVY SURF"
storm.tbl$EVTYPE[grep("CURRENT", storm.tbl$EVTYPE, ignore.case = TRUE)] <- "RIP CURRENTS/HEAVY SURF"
storm.tbl$EVTYPE[grep("HURRICANE", storm.tbl$EVTYPE, ignore.case = TRUE)] <- "HURRICANE"
storm.tbl$EVTYPE[grep("TROPICAL STORM", storm.tbl$EVTYPE, ignore.case = TRUE)] <- "TROPICAL STORM"
storm.tbl$EVTYPE[grep("SLIDE", storm.tbl$EVTYPE, ignore.case = TRUE)] <- "MUDSLIDE"
storm.tbl$EVTYPE[grep("SLUMP", storm.tbl$EVTYPE, ignore.case = TRUE)] <- "LANDSLUMP"
storm.tbl$EVTYPE[grep("DUST", storm.tbl$EVTYPE, ignore.case = TRUE)] <- "DUST STORM"
storm.tbl$EVTYPE[grep("AVALANC", storm.tbl$EVTYPE, ignore.case = TRUE)] <- "AVALANCHE"
storm.tbl$EVTYPE[grep("SURGE", storm.tbl$EVTYPE, ignore.case = TRUE)] <- "STORM SURGE/TIDE"
storm.tbl$EVTYPE[grep("WAVE", storm.tbl$EVTYPE, ignore.case = TRUE)] <- "ROGUE WAVE"
storm.tbl$EVTYPE[grep("EROS", storm.tbl$EVTYPE, ignore.case = TRUE)] <- "BEACH EROSION"
```

The above code does not clean the entire dataset. However, the goal was to clean enough of the dataset to make a representative portion reliable.

```{r}
# Calculate percentage of cleaned rows
cleanRate <- sum(head(storm.tbl %>% count(EVTYPE, sort = TRUE), 25)$n) / nrow(storm.tbl)

head(storm.tbl %>% count(EVTYPE, sort = TRUE), 20)
```

At least the top 25 most common events were categorized correctly, so at the very least, `r round(cleanRate*100, digits = 2)` percent of the dataset has a correct event type.

### Cleaning Magnitudes of Property and Agricultural Damages (PROPDMGEXP and CROPDMGEXP)

Cost of damages was unconventionally recorded in the dataset. The damage is recorded as a value (`PROPDMG` and `CROPDMG`) with a magnitude multiplier (`PROPDMGEXP` and `CROPDMGEXP`). While each event had a double stored as a value of damage cost, the magitudes were inconsistent, varying from alphabetical (K, M, B), to numeric (4, 5, 6, etc.), to symbolic (+, -, ?) to missing. 

The magnitudes were standardized as numerical powers of 10, and a singular value for each event was calculated and stored in the dataset.

```{r}
# All events with $0 in damages have magnitude of zero
storm.tbl$PROPDMGEXP[storm.tbl$PROPDMG==0] <- 0

# Change magnitudes from alphabetical to numeric
storm.tbl$PROPDMGEXP <- tolower(as.character(storm.tbl$PROPDMGEXP))
storm.tbl$PROPDMGEXP[storm.tbl$PROPDMGEXP=="h"] <- 2
storm.tbl$PROPDMGEXP[storm.tbl$PROPDMGEXP=="k"] <- 3
storm.tbl$PROPDMGEXP[storm.tbl$PROPDMGEXP=="m"] <- 6
storm.tbl$PROPDMGEXP[storm.tbl$PROPDMGEXP=="b"] <- 9
storm.tbl$PROPDMGEXP[storm.tbl$PROPDMGEXP==""] <- NA
storm.tbl$PROPDMGEXP[storm.tbl$PROPDMGEXP=="+"] <- NA
storm.tbl$PROPDMGEXP[storm.tbl$PROPDMGEXP=="-"] <- NA
storm.tbl$PROPDMGEXP <- as.numeric(storm.tbl$PROPDMGEXP)

# Calculate property damage costs
storm.tbl$PROPDMGCOST <- storm.tbl$PROPDMG * 10 ^ storm.tbl$PROPDMGEXP

missingProp <- nrow(storm.tbl[is.na(storm.tbl$PROPDMGCOST),])
```

Much of the inconsistencies were solved as the some events did not have an monetary impact (`PROPDMG = 0`), and processing the data only left `r missingProp` rows with missing data on property damages. With a large dataset of `r nrow(storm.tbl)` rows, the missing rows leave minimal impact.

Data for agricultural damages was cleaned in similar fashion. 

```{r}
# All events with $0 in damages have magnitude of zero
storm.tbl$CROPDMGEXP[storm.tbl$CROPDMG==0] <- 0

# Change magnitudes from alphabetical to numeric
storm.tbl$CROPDMGEXP <- tolower(as.character(storm.tbl$CROPDMGEXP))
storm.tbl$CROPDMGEXP[storm.tbl$CROPDMGEXP=="k"] <- 3
storm.tbl$CROPDMGEXP[storm.tbl$CROPDMGEXP=="m"] <- 6
storm.tbl$CROPDMGEXP[storm.tbl$CROPDMGEXP=="b"] <- 9
storm.tbl$CROPDMGEXP[storm.tbl$CROPDMGEXP==""] <- NA
storm.tbl$CROPDMGEXP <- as.numeric(storm.tbl$CROPDMGEXP)

# Calculate agricultural damage costs
storm.tbl$CROPDMGCOST <- storm.tbl$CROPDMG * 10 ^ storm.tbl$CROPDMGEXP

missingCrop <- nrow(storm.tbl[is.na(storm.tbl$CROPDMGCOST),])
```

Again, only `r missingCrop` rows had missing values, which is an insignificant amount. Any rows with missing damage values can be omitted from analysis.

## Results

```{r}
# Calculate total deaths/injuries
popHealthTot <- storm.tbl %>% group_by(EVTYPE) %>% 
    summarize(Deaths = sum(FATALITIES), Injuries = sum(INJURIES)) %>% 
    arrange(desc(Deaths), desc(Injuries))
```

```{r, warning=FALSE, message=FALSE}
library(reshape2)
popMelt <- melt(head(popHealthTot, 5), "EVTYPE")

library(ggplot2)
g1 <- ggplot(data=popMelt, aes(x=EVTYPE, y=value, fill=variable)) +
    geom_bar(stat="identity", position=position_dodge()) +
    labs(title = "Deaths and Injuries from Weather Events",
         x = "Event Type", y = "Total Death/Injuries (since 1955)") +
    scale_fill_discrete(name="Hazard Type")
```

```{r}
stormPropCost <- storm.tbl %>% filter(!is.na(PROPDMGCOST)) %>% 
    group_by(EVTYPE) %>% summarize(PROPAVGCOST = mean(PROPDMGCOST)) %>%
    arrange(desc(PROPAVGCOST))

stormCropCost <- storm.tbl %>% filter(!is.na(CROPDMGCOST)) %>% 
    group_by(EVTYPE) %>% summarize(CROPAVGCOST = mean(CROPDMGCOST)) %>% 
    arrange(desc(CROPAVGCOST))

stormTotCost <- stormPropCost %>% inner_join(stormCropCost, by = "EVTYPE") %>%
    mutate(TOTAVGCOST = PROPAVGCOST + CROPAVGCOST) %>% 
    arrange(desc(TOTAVGCOST))
```

```{r, warning=FALSE, message=FALSE}
p1 <- ggplot(data = head(stormPropCost, 5), aes(x=EVTYPE, y=PROPAVGCOST, fill=EVTYPE)) + 
    geom_bar(stat = "identity") + guides(fill=FALSE) + 
    labs(title = "Property Damages from Weather Events",
         x = "Event Type", y = "Cost of Damages (in US Dollars)")

p2 <- ggplot(data = head(stormCropCost, 5), aes(x=EVTYPE, y=CROPAVGCOST, fill=EVTYPE)) + 
    geom_bar(stat = "identity") + guides(fill=FALSE) + 
    labs(title = "Agricultural Damages from Weather Events",
         x = "Event Type", y = "Cost of Damages (in US Dollars)")

library(grid)
library(gridExtra)
g2 <- grid.arrange(p1, p2, ncol=2)
```

```{r}
stormCostMelt <- melt(head(stormTotCost[,1:3], 3), "EVTYPE")

g3 <- ggplot(data=head(stormTotCost, 5), aes(x=EVTYPE, y=TOTAVGCOST, fill=EVTYPE)) +
    geom_bar(stat = "identity") + guides(fill=FALSE) +
    labs(title = "Economical Consequences of Weather Events",
         x = "Storm Type", y = "Total Cost of Damages (in US Dollars)")